---
title: "HW3-drafting-viz"
author: "Amanda Overbye"
date: 02/12/2025
toc: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  eval: true
  warning: false
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, code_folding = FALSE)
```

# Questions & Answers

**1. Which option do you plan to pursue?**

**2. Restate your question(s). Has this changed at all since HW #1? If
yes, how so?**

**3. Explain which variables from your data set(s) you will use to
answer your question(s), and how.**

**4. Find at least two data visualizations that you could (potentially)
borrow / adapt pieces from. Link to them or download and embed them into
your .qmd file, and explain which elements you might borrow (e.g. the
graphic form, legend design, layout, etc.)**

**5. Hand-draw your anticipated visualizations, then take a photo of
your drawing(s) and embed it in your rendered .qmd file – note that
these are not exploratory visualizations, but rather your plan for your
final visualizations that you will eventually polish and submit with**

**6 Mock up all of your hand drawn visualizations using code.**

**7. Answer the following questions:**

a.  *What challenges did you encounter or anticipate encountering as you
    continue to build / iterate on your visualizations in R? If you
    struggled with mocking up any of your three visualizations (from #6,
    above), describe those challenges here.*
b.  *What ggplot extension tools / packages do you need to use to build
    your visualizations? Are there any that we haven’t covered in class
    that you’ll be learning how to use for your visualizations?*
c.  *What feedback do you need from the instructional team and / or your
    peers to ensure that your intended message is clear?*

## Load necessary packages and read in your data.

```{r}
# Load packages
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(tmap)
library(here)
library(janitor)
library(usdata)
library(tigris)
library(sf)
library(scales)
library(airportr)
library(ggmap)
library(maps)
```

```{r}
# Testing maps
map_data_es <- map_data('world')[map_data('world')$region == "USA",]
world <- worldMapEnv
```

#### Link to data

https://data.transportation.gov/Aviation/International_Report_Freight/u4sg-r5vg/about_data

*More information*

https://www.transtats.bts.gov/freight.asp

```{r}
# Load in the data
freight2024 <- read.csv(here("data", "flights2024.csv")) %>% 
  filter(FREIGHT != 0)

air_coords <- airports
```

```{r}
# Function for reading in CSVs
read <- function(year) {
  file_name <- paste0("flights", year, ".csv") 
  file_path <- here("data", file_name) 
  
  # Read and filter dataset
  df <- read.csv(file_path) %>%
    filter(FREIGHT != 0)
  
  return(df)
}
```

```{r}
freight2023 <- read(2023)
```

# Hand-drawn anticipated visualizations

*a sketch of your infographic (which should include at least three
component visualizations) if you are pursuing option 1 a sketch of all
three separate visualizations if you are pursuing option 2 a sketch of
the approved number of visualizations, as outlined in your proposal, if
you are pursuing option 3*

# Mock up of hand drawn visualizations using code
