---
title: "HW3-drafting-viz"
author: "Amanda Overbye"
date: 02/12/2025
toc: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  eval: true
  warning: false
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, code_folding = FALSE)
```

# Questions & Answers

**1. Which option do you plan to pursue?**

**2. Restate your question(s). Has this changed at all since HW #1? If
yes, how so?**

**3. Explain which variables from your data set(s) you will use to
answer your question(s), and how.**

**4. Find at least two data visualizations that you could (potentially)
borrow / adapt pieces from. Link to them or download and embed them into
your .qmd file, and explain which elements you might borrow (e.g. the
graphic form, legend design, layout, etc.)**

**5. Hand-draw your anticipated visualizations, then take a photo of
your drawing(s) and embed it in your rendered .qmd file – note that
these are not exploratory visualizations, but rather your plan for your
final visualizations that you will eventually polish and submit with**

**6 Mock up all of your hand drawn visualizations using code.**

**7. Answer the following questions:**

a.  *What challenges did you encounter or anticipate encountering as you
    continue to build / iterate on your visualizations in R? If you
    struggled with mocking up any of your three visualizations (from #6,
    above), describe those challenges here.*
b.  *What ggplot extension tools / packages do you need to use to build
    your visualizations? Are there any that we haven’t covered in class
    that you’ll be learning how to use for your visualizations?*
c.  *What feedback do you need from the instructional team and / or your
    peers to ensure that your intended message is clear?*

## Load necessary packages and read in your data.

```{r}
# Load packages
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(tmap)
library(here)
library(janitor)
library(usdata)
library(tigris)
library(sf)
library(scales)
library(airportr)
library(ggmap)
library(maps)
library(patchwork) 
```

```{r}
# Testing maps
map_data_es <- map_data('world')[map_data('world')$region == "USA",]
world <- worldMapEnv
```

#### Link to data

https://data.transportation.gov/Aviation/International_Report_Freight/u4sg-r5vg/about_data

*More information*

https://www.transtats.bts.gov/freight.asp

```{r}
# Load in the data
freight2024 <- read.csv(here("data", "flights2024.csv")) %>% 
  filter(FREIGHT != 0)

air_coords <- airports
```

```{r}
# Function for reading in CSVs
read <- function(year) {
  file_name <- paste0("flights", year, ".csv") 
  file_path <- here("data", file_name) 
  
  # Read and filter dataset
  df <- read.csv(file_path) %>%
    filter(FREIGHT != 0) %>% 
    clean_names()
  
  return(df)
}
```


```{r}
# Function for reading in CSVs
test <- function(year) {
  file_name <- paste0("flights", year, ".csv") 
  file_path <- here("data", file_name) 
  
  # Read and filter dataset
  df <- read.csv(file_path) %>%
    filter(FREIGHT != 0) %>%
    
    clean_names() %>% 
    
  return(df)
}
```


```{r}
freight2023 <- read(2023)
```

```{r}
freight2023 <- read(2023)
freight2022 <- read(2022)
freight2021 <- read(2021)
freight2020 <- read(2020)
freight2019 <- read(2019)
freight2018 <- read(2018)
freight2017 <- read(2017)
freight2016 <- read(2016)
freight2015 <- read(2015)
freight2014 <- read(2014)
```

clean_counties <- clean_counties %>% 
  filter(!state %in% c("American Samoa"

```{r}
# top  10 questions
top_10 <- freight2023 %>%
  filter(!origin_country_name %in% "United States") %>% 
  count(origin_country_name, sort = TRUE) %>% 
  slice_max(n, n = 10) %>% 
  arrange(desc(n)) 
```

### Plot Countries

```{r}
country_plot <- ggplot(top_10, aes(x = fct_reorder(origin_country_name,desc(n)),
                   y = n, 
                   fill = origin_country_name)) +
  geom_col() +
  theme_void() +
  theme(legend.position = "none")
```

### Plot timeline

```{r}
# TEST
# Step 1: Find the top 5 countries with the highest yearly flight count
top_5_countries <- freight2023 %>%
  filter(!origin_country_name %in% "United States") %>% 
  count(origin_country_name, sort = TRUE) %>%  # Count total flights per country
  slice_max(n, n = 5) %>%  # Select top 5 countries
  pull(origin_country_name)  # Extract country names as a vector

# Step 2: Get monthly flight counts for only the top 5 countries
top_5_monthly <- freight2023 %>%
  filter(origin_country_name %in% top_5_countries) %>%  # Keep only top 5 countries
  count(origin_country_name, month) %>%  # Count flights per month
  arrange(origin_country_name, month) %>%   # Sort results
pivot_wider(names_from = origin_country_name, values_from = n, values_fill = 0) %>%  
  arrange(month)
```

The plotting

```{r}
timeline <- ggplot(top_5_monthly, aes(month)) + 
  geom_line(aes(y = Mexico, colour = "Mexico")) +
  geom_line(aes(y = Japan, colour = "Japan")) +
  geom_line(aes(y = Canada, colour = "Canada")) +
  geom_line(aes(y = Germany, colour = "Germany"))+
  labs(title = "Tons of Domestic Freigh Flight Metrics compared to international",
    subtitle = "air freight metric traveling between international points and U.S. airports.") +
    theme_void()
```



# Hand-drawn anticipated visualizations

*a sketch of your infographic (which should include at least three
component visualizations) if you are pursuing option 1 a sketch of all
three separate visualizations if you are pursuing option 2 a sketch of
the approved number of visualizations, as outlined in your proposal, if
you are pursuing option 3*

# Mock up of hand drawn visualizations using code

## Create base plot
```{r}
plot_base <- ggplot() +
  
  # Add labels, subtitles, and caption to plot
  labs(
    title = "Airplanes & Freight",
    subtitle = "Airplanes carry a lot of freight different places",
    caption = "they can flY!"
    ) +
  
  # apply a completely empty theme ----
  theme_void() +
  
  # further customize theme ----
  theme(
    
   # Set preferences for text including font and size
    text = element_text(size = 16, 
                        lineheight = 0.3, 
                        color = "black"),
    
    # Fill in the background with color
    
    # make plot background black
    plot.background = element_rect(fill = "lightblue1", 
                                   color = "lightblue1"),
    
    # Plot the title and add text preferences
    plot.title = element_text(size = 42, 
                              face = "bold", 
                              hjust = 0.5, 
                              margin = margin(b = 10)),
    
    # Plot subtitle and add text preferences
    plot.subtitle = element_text(hjust = 0.5, 
                                 margin = margin(b = 20)),
    
    # format plot caption using {ggtext} to render fontawesome icons ----
    plot.caption = ggtext::element_markdown(face = "italic",
                                            color = "white",
                                            hjust = 0.5,
                                            margin = margin(t = 20)),

    # Add margins and space to the plot
    plot.margin = margin(b = 20, t = 50, r = 50, l = 50)
    
  )
```

```{r}
plot_base
```

# Add elements together!

```{r}
plot_final <- plot_base +
  
  # Lay out the order plots will be placed on the graph and set their coordinates
  inset_element(country_plot, left = 0, right = .5, top = .5, bottom = 0)+
  inset_element(timeline, left = 0, right = .5, top = 1, bottom = .5)
```

```{r}
plot_final
```

