---
title: "The Bump Chart"
author: "Amanda Overbye"
date: 02/12/2025
toc: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
execute: 
  eval: true
  warning: false
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, code_folding = FALSE)
```

# Load Libraries and Data

```{r}
# Load packages
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(tmap)
library(here)
library(janitor)
library(usdata)
library(tigris)
library(sf)
library(scales)
library(airportr)
library(ggmap)
library(maps)
library(patchwork) 
library(lubridate)
library(gridExtra)
library(ggalluvial)
library(geosphere)
library(waffle)
library(showtext)
library(ggbump)
```

### Function for reading in data

```{r}
# Function for reading in CSVs
read <- function(year) {
  file_name <- paste0("flights", year, ".csv") 
  file_path <- here("data", file_name) 
  
  # Read and filter dataset
  df <- read.csv(file_path) %>%
    filter(FREIGHT != 0) %>% 
    clean_names()
  
  return(df)
}
```

### Reading in the data via the function

```{r}
# Read in data via function

freight2024 <- read(2024)
freight2023 <- read(2023)
freight2022 <- read(2022)
freight2021 <- read(2021)
freight2020 <- read(2020)
freight2019 <- read(2019)
freight2018 <- read(2018)
freight2017 <- read(2017)
freight2016 <- read(2016)
freight2015 <- read(2015)
freight2014 <- read(2014)
freight2013 <- read(2013)
freight2012 <- read(2012)
freight2011 <- read(2011)
freight2010 <- read(2010)
freight2009 <- read(2009)
freight2008 <- read(2008)
freight2007 <- read(2007)
freight2006 <- read(2006)
freight2005 <- read(2005)
freight2004 <- read(2004)
```

### Function to join dfs

```{r}
# Create function to join datasets
join_freight_data <- function(...) {
  Reduce(full_join, list(...))
}
```

### Joining the dfs via function

```{r}
# Use the function to join all datasets
freight_all <- join_freight_data(
  freight2004, freight2005, freight2006, freight2007, 
  freight2008, freight2009, freight2010, freight2011, 
  freight2012, freight2013, freight2014, freight2015, 
  freight2016, freight2017, freight2018, freight2019, 
  freight2020, freight2021, freight2022, freight2023, 
  freight2024
)
```

# Data Wrangling for Bump Chart

# COPY

```{r}
#...................rank occupations by salary...................
country_rank_by_year <- freight_all |> 
  select(year, occupation, total_earnings) |> 
  group_by(year) |> 
  mutate(
    rank = row_number(desc(total_earnings))
  ) |> 
  ungroup() |> 
  arrange(rank, year)

#........get top 8 occupation names for final year (2016)........
top2016 <- salary_rank_by_year |>  
  filter(year == 2016, rank <= 8) |>  
  pull(occupation) 

```

## color copy 

```{r}
magma_pal <- viridisLite::magma(12)
```

# COPY

**Note to self to help thinking** 
Because all of this data has to do with the USA, each row will have the US in either dest or origin. So, to get the countries with the most amount of freight, it will be useful to make a column with the country involved in each row that is not the united states. 

```{r}
# Get the amount of freight based on the non us country 
other_countries_col <- freight_all %>% 
  mutate(other_country = ifelse(origin_country_name == "United States", dest_country_name, origin_country_name)) %>% 
  group_by(other_country, year) %>%
  summarize(total_freight = sum(freight, na.rm = TRUE))
```

Filtered it to after 2019, so lines didnt go outside of chart

```{r}
# Rank the number one countries by year
country_rank_by_year <- other_countries_col %>%
  filter(year >= 2019) %>% 
  select(year, other_country, total_freight) %>% 
  group_by(year) %>% 
  mutate(
    rank = row_number(desc(total_freight))
  ) %>% 
  ungroup() %>% 
  arrange(rank, year)
```

### Filter to get years of interest

```{r}
# Filter for 2024
top_countries_2024 <- country_rank_by_year %>% 
  filter(year == 2024, rank <= 10) %>% 
  pull(other_country)

# Filter for 2019
#top_countries_2019 <- country_rank_by_year %>% 
#filter(year == 2019, rank <= 10)
```

# Create bump chart


```{r}
country_rank_by_year %>%
  filter(other_country %in% top_countries_2024) %>%
  ggplot(aes(x = year, y = rank, color = other_country)) + 
  geom_point(shape = "|", size = 6) + 
  geom_bump(linewidth = 1) +
  geom_text(
    data = country_rank_by_year %>% 
      filter(year == 2019, other_country %in% top_countries_2024),
    aes(label = other_country),
    hjust = 1,
    nudge_x = -0.1
  ) +
  geom_text(
    data = country_rank_by_year %>%  
      filter(year == 2024, other_country %in% top_countries_2024),
    aes(label = rank),
    hjust = 0,
    nudge_x = 0.1
  )  +
  annotate(
    geom = "text",
    x = c(2019, 2024),
    y = c(-0.2, -0.2),
    label = c("2019", "2024"),
    hjust = c(0, 1),
    vjust = 0,
    size = 5) +
  scale_y_reverse() +
  scale_color_manual(values = magma_pal, guide = "none") +
  coord_cartesian(xlim = c(2019, 2024), #make the plot smaller 
                  ylim = c(11, 0.25), 
                  clip = "off") +
  theme(
    axis.title = element_text(size = 12),
    axis.text.y=element_blank(),
    legend.position = "none"
  ) +
  theme_void()

```



